Sufficient Conditions for Reachability in Automata Networks with Priorities

### 2014-12-21 main.tex
more corrections from the suggestions of the reviewers

In this paper, we propose an extension of the Process Hitting framework
by using classes of priorities in order to improve its expressivity.
We also show that these priorities allow to represent accurate Boolean gates in order to model
cooperations between components
(that is, for instance, when two components are supposed to interact with a third one
only when they are in a given configuration)
which was not possible before.

Furthermore, we develop in this paper a new refinement for the under-approximation of the static analysis
to give accurate results on Process Hitting models.
This method applies to a specific sub-class of models,
but this class turns out to be sufficiently expressive to represent any
Process Hitting model with an arbitrary number of classes of priorities,
thus guaranteeing a wide field of application of our method.
Finally, we show that Process Hitting allows to accurately model
any Asynchronous Discrete Networks, either Boolean or multivalued.
Our static analysis method thus allows to efficiently under-approximate reachability properties
in Asynchronous Discrete Networks,
as it is illustrated on the model-checking of a signalling network of 94
components, which is unprecedented.
\end{abstract}
\begin{keyword}
discrete networks \sep
abstract interpretation \sep
reachability \sep
qualitative models \sep
systems biology
\end{keyword}
\end{frontmatter}

>>>>

In this paper, we propose an extension of the Process Hitting framework
by using classes of priorities in order to improve its expressivity.
Furthermore, we develop a new refinement for the under-approximation of the static analysis
to give accurate results on Process Hitting models.
This method applies to a specific sub-class of models,
but this class turns out to be sufficiently expressive to represent any
Process Hitting model with an arbitrary number of classes of priorities,
thus guaranteeing a wide field of application of our method.

Finally, we also show that the priorities
allow to represent accurate Boolean gates in order to model
cooperations between components
(that is, for instance, when two components are supposed to interact with a third one
only when they are in a given configuration)
which was not possible before.
This allows to accurately model
Asynchronous Discrete Networks, either Boolean or multivalued,
and efficiently under-approximate reachability properties
in these models.
We illustrate this on the model-checking of a signalling network of 94
components, which is unprecedented.
\end{abstract}
\begin{keyword}
discrete networks \sep
abstract interpretation \sep
reachability \sep
qualitative models \sep
systems biology
\end{keyword}
\end{frontmatter}

### 2014-12-21 parts/intro.tex
more corrections from the suggestions of the reviewers

Of course, such formal models on which analyses are performed are abstract representations of the actual studied systems.
They are associated with parameters that have to be synthesised %so as to be as much as possible fitting with the real systems having some observed behaviors.
to give the most faithful representation of the real systems with their observed behaviours.
As a matter of fact, the abstractions we get are more or less rough or accurate.
Prevalent formal frameworks for such modelling activities are state-transition systems or process algebras. % Petri nets
We developed a quite similar framework named the Process Hitting \cite{PMR10-TCSB},
consisting in a restriction of these frameworks where the evolution of a component is determined by the state of at most one other component that does not evolve.
In a sense, these kind of actions are of the form $X + Y \rightarrow X + Y'$ where $X$ behaves like a catalyst molecule that “hits” another molecule $Y$ and changes it into $Y'$, without being changed itself.
Assuming catalysts are always available, this can represent any biochemical system made of monomolecular reactions, and can also represent catalytic networks such as metabolic networks.
%, with the aim of avoiding to build the whole state space in order to be able to tackle very large systems
Our motivation behind this framework was to design a model and analysis techniques adapted to biological modelling.
These analyses avoid to build the whole state space, which allows to tackle very large systems (that would have led to a huge number of states, hopelessly too huge to be analysed).
They are based on the fact that most biological models have few levels of expression per component:
in Boolean networks~\cite{kauffman69,Thomas73} there are only two levels per component and in their multivalued equivalent, Asynchronous Discrete Networks~\cite{deJong02}, components rarely have more than four levels.

@@

Whereas the analysis presented in \pref{sec:sa} considers a particular class of
Process Hitting with two classes of priority,
we provide in \pref{sec:encodings} a “flattening” of any Process Hitting model into
this class of models,
so that its application is not restricted to it.
We also give in the same section a formal encodings of Asynchronous Discrete Networks,
allowing to efficiently study this kind of widespread models.

>>>>

Of course, such formal models on which analyses are performed are abstract representations of the actual studied systems.
They are associated with parameters that have to be synthesised %so as to be as much as possible fitting with the real systems having some observed behaviors.
to give the most faithful representation of the real systems with their observed behaviours.
As a matter of fact, the abstractions we get are more or less rough or accurate.
Prevalent formal frameworks for such modelling activities are state-transition systems or process algebras. % Petri nets
We developed a quite similar framework named the Process Hitting \cite{PMR10-TCSB},
consisting in a restriction of these frameworks where the evolution of a component is determined by the state of at most one other component that does not evolve.
In a sense, these kind of actions are of the form $X + Y \rightarrow X + Y'$ where $X$ behaves like a catalyst molecule that “hits” another molecule $Y$ and changes it into $Y'$, without being changed itself.
Assuming catalysts are always available, this can represent any biochemical system made of monomolecular reactions, and can also represent catalytic networks such as metabolic networks.
%, with the aim of avoiding to build the whole state space in order to be able to tackle very large systems
Our motivation behind this framework was to design a model and analysis techniques adapted to biological modelling.
These analyses avoid to build the whole state space, which allows to tackle very large systems (that would have led to a huge number of states, hopelessly too huge to be analysed).
They are based on the fact that most biological models have few levels of expression per component:
in Boolean networks~\cite{kauffman69,Thomas73} there are only two levels per component, and in their multivalued equivalent~\cite{deJong02}, components rarely have more than four levels.

@@

Whereas the analysis presented in \pref{sec:sa} considers a particular class of
Process Hitting with two classes of priority,
we provide in \pref{sec:encodings} a “flattening” of any Process Hitting model into
this class of models,
so that its application is not restricted to it.
We also give in the same section a formal encoding of Asynchronous Discrete Networks,
that is, the multivalued version of Asynchronous Boolean Networks.
These models,
which encompass variables and evolution functions with a limited number of discrete values,
also called “Logical Networks” in~\cite{Thomas95,deJong02}.
This encoding thus allows to efficiently study this kind of widespread models.

### 2014-12-21 parts/models-ph-dn.tex
more corrections from the suggestions of the reviewers

We exhibit an encoding of Asynchronous Discrete Networks (ADN) with the Process
Hitting using two classes of priorities
and prove a weak bisimulation relation.
This translation is important as it allows to efficiently study the dynamics ADNs
by using the static analysis developed in \pref{sec:sa};
indeed, the bisimulation relation ensures that the dynamics of the
resulting PH model are accurately the same than the original.
Furthermore, the obtained model is ensured to respect
all Conditions of \pref{ssec:hypothesis}.

>>>>

We exhibit in this section an encoding of Asynchronous Discrete Networks (ADN),
also called Logical Networks or equivalently René Thomas Models \cite{Thomas95,deJong02},
with the Process Hitting using two classes of priorities
and prove a weak bisimulation relation.
This translation is important as it allows to efficiently study the dynamics ADNs
by using the static analysis developed in \pref{sec:sa};
indeed, the bisimulation relation ensures that the dynamics of the
resulting PH model are accurately the same than the original.
Furthermore, the obtained model is ensured to respect
all Conditions of \pref{ssec:hypothesis}.
